---
title: "PUG Blog Project"
authors: Maggie Smith & Sammy Phillips
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggrepel)
library(GGally)
library(ggmosaic)
```

# COVID-19 Impact on Individual Economic Health

This project explores the impact of COVID-19 on individual economic health. We used data collected by the COVID Impact Survey. The COVID Impact Survey was an independent, non-governmental survey carried out by the Data Foundation in late spring of 2020 focusing on physical health, economic and food security, and employment. The data is publicly available in a csv format: https://www.covid-impact.org/results.

First, we will determine how respondents of the COVID Impact Survey group together based on various demographic and economic health questions using unsupervised learning and clustering. Then, we will examine the relationships between other demographic and economic health variables and cluster assignment.

# Part 1: Clustering Analysis

First, we perform kmeans clustering analysis with 2 clusters, 4, clusters, 7 clusters, and 11 clusters.

Our clustering analysis includes the following variables:
* AGE7 - Age: 7 category
* HHINCOME - Household income
* EDUCATION - What is the highest level of school you have completed?
* ECON4A - Think about 30 days from now, how likely do you think it is that you will be employed at that time?
* ECON4B - Think about 3 months from now, how likely do you think it is that you will be employed at that time?

```{r}
kmeans_data <- read.csv("~/Desktop/DataScience/git/Blog-Girls-Who-Code/kmeans_data.csv") 

set.seed(23)
km2_out <- kmeans(kmeans_data, centers = 2, nstart = 20)
km2_out$centers

kmeans_data2 <- kmeans_data %>%
  mutate(clust2 = as.character(km2_out$cluster))

set.seed(23)
km4_out <- kmeans(kmeans_data, centers = 4, nstart = 20)
km4_out$centers

kmeans_data2 <- kmeans_data2 %>%
  mutate(clust4 = as.character(km4_out$cluster))

set.seed(23)
km5_out <- kmeans(kmeans_data, centers = 5, nstart = 20)
km5_out$centers

kmeans_data2 <- kmeans_data2 %>%
  mutate(clust5 = as.character(km5_out$cluster))

set.seed(23)
km7_out <- kmeans(kmeans_data, centers = 7, nstart = 20)
km7_out$centers

kmeans_data2 <- kmeans_data2 %>%
  mutate(clust7 = as.character(km7_out$cluster))

set.seed(23)
km11_out <- kmeans(kmeans_data, centers = 11, nstart = 20)
km11_out$centers

kmeans_data2 <- kmeans_data2 %>%
  mutate(clust11 = as.character(km11_out$cluster))

```

Next, we determine which clustering analysis is best. We do this by creating an elbow plot to determine how many clusters are needed to produce a small amount of total with-in cluster variation. This plot does not display a definitive elbow. However, the decrease in total with-in cluster variation begins to get smaller after 4 clusters. For this reason, our analysis with 4 clusters is likely our best kmeans clustering analysis. We will use this analysis for the rest of our investigation.

```{r}
fig <- matrix(NA, nrow=15, ncol=2)

set.seed(75)
for (i in 1:15){
  fig[i,1] <- i
  fig[i,2] <- kmeans(kmeans_data
                    , centers=i
                    , nstart=20)$tot.withinss
}

ggplot(data = as.data.frame(fig), aes(x = V1, y = V2)) +
  geom_point() + 
  geom_line() +
  scale_x_continuous(breaks=c(1:15)) +
  labs(x = "K", y = expression("Total W"[k]))
```

Finally, we can visualize our kmeans clustering analysis with 4 clusters below.

```{r}
vars <- c("ECON4A", "ECON4B", "HHINCOME", "EDUCATION", "AGE7")

ggpairs(data = kmeans_data2
        , aes(color = clust4) 
        , columns = vars
        , upper = list(continuous = "blank"))
```

# Part 2: Post-Clustering Analysis

Now, we want to explore the relationship between cluster assignment and a number of different demographic and economic health variables. 

The demographic variables of interest are: 
* GENDER - Are you male or female?
* RACETH - Race/ethnicity
* P_DENSE - Population density

The economic health variables of interest are:
* ECON6A - [Unemployment insurance] In the past 7 days, have you either received, applied for, or tried to apply for any of the following forms of income or assistance, or not?
* ECON6C - [TANF (Temporary Assistance for Needy Families)] In the past 7 days, have you either received, applied for, or tried to apply for any of the following forms of income or assistance, or not?
* ECON7_1 - [Put it on my credit card and pay it off in full at the next statement] Based on your current financial situation, how would you pay for this expense?
* ECON7_2 - [Put it on my credit card and pay it off over time] Based on your current financial situation, how would you pay for this expense?
* ECON7_3 - [Use money currently in my checking or savings account or with cash] Based on your current financial situation, how would you pay for this expense?
* ECON7_4 - [Use money from a bank loan or line of credit] Based on your current financial situation, how would you pay for this expense?
* ECON7_5 - [Borrow from a friend or family member] Based on your current financial situation, how would you pay for this expense?
* ECON7_6 - [Use a payday loan, deposit advance or overdraft] Based on your current financial situation, how would you pay for this expense?
* ECON7_7 - [Sell something] Based on your current financial situation, how would you pay for this expense?
* ECON7_8 - [I would not be able to pay for it right now] Based on your current financial situation, how would you pay for this expense?
* ECON5A_B - [The food that we bought just didn't last, and we didn't have money to get more] Please indicate whether the following statements were often true, sometimes true, or never true for you or your household over the past 30 days.

First, we create visualizations to explore the relationship between cluster assignment and our demographic variables of interest. 

```{r}
post_cluster_data <- read.csv("~/Desktop/DataScience/git/Blog-Girls-Who-Code/post_cluster_data.csv") %>% 
  mutate(clust4 = as.character(km4_out$cluster))

ggplot(post_cluster_data, aes(x = clust4, fill = GENDER2)) +
  geom_histogram(stat = "count")

ggplot(post_cluster_data, aes(x = clust4, fill = RACETH2)) +
  geom_histogram(stat = "count")

```


